---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

1.  Discuss and navigate different data types in R
2.  Create, manipulate, and explore datasets
3.  Date objects

## Data Types in R

R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type.

-   **Matrices** 2 dimensional structure that contains elements of the same type.

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes.

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (\# at beginning of line, ---- at end)

```{r Create various data structures}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # 

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # 
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) # if you choose you can organize 'bycolumn' which'll change how it organizes your data in your matrix
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "quatro", "cinco"), 
                                  c("un", "deux", "trois", "quatre"))) #

matrix1[4, ] # whatever comes before the comma is usually wrt to the row of the matrix and after row would be the column. when you don't specify you get the entire row
matrix1[ , 3] #
matrix1[c(12, 14)] # this means you want to point to two elements, element 12 and element 14 - this will start counting by row until it hits element 12
matrix1[c(12:14)] #
matrix1[2:4, 1:3] # if you want a subset of rows and a subset of columns, you're saying rows 2, 3, and 4 and columns 1, 2, 3. default will always be 1 unless you specify a seq

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]]

# Data Frames ---- # how to rename data frames
d <- c(1, 2, 3, 4) # What type of vector?
e <- c("red", "white", "red", NA) # What type of vector?
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector?
dataframe1 <- data.frame(d,e,f) # 
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # the argument will be the data frame ur trying to rename and then a column or a vector represented by a little c and the strings we're trying to assign to the data frame. D E F were replaced by ID Color and Passed. Always inspect your data frame, always check your environment to check everything's going smoothly and is where it needs to be

dataframe1[1:2,] # 
dataframe1[c("ID","Passed")] # 
dataframe1$ID

```

Question: How do the different types of data appear in the Environment tab?

> Answer:

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer:

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r DataExercise-1a}
# manually create those two vectors, high and low
f_high <- as.vector(c(81,82,83,84,79,71,78,82,85,84))
f_high
f_low <- as.vector(c(55,56,58,60,60,62,62,64,64,62))
f_low

```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r DataExercise-1b}
# convert to Celsius and the formula is Fahrenheit - 32 x 5/9 = Celsius
# now you have four vectors, two high and lows for diff data units

c_high <- (f_high - 32)*(5/9)
c_high
c_low <- (f_low - 32)*(5/9)
c_low
#ALTERNATIVE WAY
convert_to_celsius <- function(x){
  (x-32)*(5/9)
  x
}
convert_to_celsius

c_high <- convert_to_celsius(f_high)
print(c_high)
c_low <- convert_to_celsius(f_low)
print(c_low)
```

Combine your four vectors into a data frame and add informative column names.

```{r DataExercise-1c}
df_temp <- data.frame("HighTempF"=f_high, "HighTempC"=c_high, "LowTempF"=f_low, "LowTempC"=c_low)
df_temp

```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r DataExercise-1d}
# function summary to get mean and function sd need to specify the specific column where we want the standard dev to be calculated
summary(df_temp)
# sd(df_temp) gives you an error bc sd doesn't accept data frames as an argument so u need to give it an object
sd(df_temp$HighTempF) #sd will work over matrices or anything with all the same class 
```

### Date objects

Remember formatting of dates in R:

%d day as number (0-31) %m month (00-12, can be e.g., 01 or 1) %y 2-digit year %Y 4-digit year %a abbreviated weekday %A unabbreviated weekday %b abbreviated month %B unabbreviated month

```{r Coding with dates}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented)
today <- Sys.Date()
#format(today, format = "%B")
format(today, format = "%d - %m - %y")
#format(today, format = "")
#format(today, format = "")

```

### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][<https://cran.r-project.org/web/packages/lubridate/lubridate.pdf>].

```{r DateExercise1}
#install.packages("lubridate")
library(lubridate)
#need to parse factors into date objects then you'll be able to use dates 

#Ex1
str_today <- "2023-feb-7"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today) #call fxn ymd bc that's the order yr month date and the only object it'll take is str_today
date_obj_today

#Ex2
str_today2 <- "Feb 7, 2023"
#Since the format is month-day-year we will use function mdy()
date_obj_today <- mdy(str_today2)
date_obj_today

#Ex_3 - on your own...
str_juneteenth <- "19 June 1865"
#Since the format is month-day-year we will use function mdy()
date_juneteenth <- dmy(str_juneteenth)
date_juneteenth
```

```{r DateExercise2}
#century issue - whenever ur data set goes too far back in time and it's only expressed in two digits you might get some issues with the conversions, IT'S ONLY A PROBLEM IF YOUR DATES GO BACK TO BEFORE 1970 JAN 1
str_past <- "55-feb-3"
date_obj_past <- ymd(str_past)
date_obj_past

#Build a function to fix year that is more general than the one discussed in the lesson -- a fxn u can run to fix the issue above.
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d)
}

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=23) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue
str_past <- "55-feb-3"
#Alternative 1
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=23L)
date_obj_past

#Alternative 2
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=23L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine.

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin

```
