---
title: "5: Lab - Data Visualization Basics"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M4/A4
2. Answer questions on M5 videos
3. Additional comments on videos - expression(), saving plots, factors
4. Perform simple data visualizations in the R package `ggplot`


## Load packages, etc.

```{r, message = FALSE, warning = FALSE}
#Import basic libraries
library(tidyverse);library(lubridate);library(here)

#install.packages(colormap)

#Library for creating ridge plots
library(ggridges)

#Enhanced color ramps
library(viridis)
library(RColorBrewer)
library(colormap)

#Add pre-set themes <--NEW
library(ggthemes)
```

## Expression function and Saving plots

The expression function can be used when you want to add mathematical symbols and subscripts mixed with regular letter to labels. 

```{r formatted text}

#Create a plot
Plot_text <-
ggplot(data=data.frame(x=0,y=0))+
  geom_point(aes(x=x,y=y)) +
  geom_text(
    x=0,
    y=0.025,
    # bc we have the expression function below, r know you're asking for the greek letter mu -- it will transform it into an equation
    #if u want an r squared value for linear regression. if you use the carrot ^, r will know you're taking the thing to the power of something. Just like you'd add expressions to a word, it doesn't matter if you use double or single quotes. it's still a character
    label=expression(paste("Mean is ", mu,",", R^{2},'=0.6'))
    )

#Print the plot - reason you got mu and r squared like that is bc i used the expression function. if you have expressions, you need to make sure you use the expression function so they get converted to an equation
print(Plot_text)

#Save the plot - specifying the path and the name of the file. can change the height and width of your plot below but it will only change how it shows up on the actual knitted file / exported file
ggsave(
  filename = "./Output/Plot_text.jpg",
  plot = Plot_text, 
  height = 4, 
  width = 6, 
  units = "in", 
  dpi = 300
  )

```

---

## Import Datasets
*Note the use of a string variable the points to the location of the data combined with the `here()` package functionality. This enables us to quickly change where we get our data, i.e., by changing one line of code vs each `read.csv()` command.* 

```{r, message = FALSE, warning = FALSE}
#Assign a variable to the processed data folder location
processed_data = "Data/Processed_KEY" 

#Read in data
PeterPaul.chem.nutrients <- read.csv(
  here(processed_data,"NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

EPAair <- read.csv(
  here(processed_data,"EPAair_O3_PM25_NC1819_Processed.csv"), 
  stringsAsFactors = TRUE)

#Fix dates
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
EPAair$Date <- ymd(EPAair$Date)

```

## A note on factors in plots...
Say we want to plot the number of records each month in the Peter/Paul lakes dataset. How might we do that? What challenges do we face? 

```{r Plot the number of records by month}
#What kind of values are months?
class(PeterPaul.chem.nutrients$month)

#List unique values
unique(PeterPaul.chem.nutrients$month)

#Plot how many Lake observations occur each month
ggplot(PeterPaul.chem.nutrients,
       aes(x=month))+
  geom_bar()

#Convert month to a factor -- with 12 levels, labelled with month names (month.abb give you twelve months of yr w abbreviated format - now instead of numbers and integers you have name of month representing each instance. didn't chance orig data set, just applied function to that specific column -- can always do this nested within ggplot
factor(PeterPaul.chem.nutrients$month,levels=1:12,labels=month.abb)

#Try again: Plot how many Lake observations occur each month 
ggplot(PeterPaul.chem.nutrients,
       aes(x=factor(month,levels=1:12, labels=month.abb)))+
  geom_bar()

#Alter to include missing months -- need to add something to the x axis to make sure it's not dropping any levels. the scalex u should use will depend on the class (continuous variable? go with x. discrete? go w discrete. date? go with date. our month is not a date object in this case so it should not be date. iin this case, use discrete bc we defined month as a factor, which is just a category, so discrete values)
ggplot(PeterPaul.chem.nutrients,
       aes(x=factor(month,levels=1:12, labels=month.abb)))+
  geom_bar()
#name = month renames x axis. if you dont wanna drop any level, include drop=FALSE
  scale_x_discrete(name="",drop=FALSE)
  #to put number values above the columns, you need to add labels that are equal to the y values
  

#Tidy up the code (try to avoid writing a lot of things next to each other, so every time you have a comma try hitting return to go to the next line)

```


---

## More on themes

Themes are used to control the overall appearance of a plot, such as the color, size, font, and style of the various elements in the plot. Themes provide a quick and easy way to change the visual style of a plot, and they allow us to create custom visualizations that fit our specific needs and preferences. The reason we are taught ggplot is bc of all the flexibility.

### Themes control the following elements
 * _Plot background_: The background color or fill pattern of the plot area.
 * _Plot title_: The size, font, and position of the plot title.
 * _Axis labels_: The font, size, and position of the x-axis and y-axis labels.
 * _Axis ticks and grid lines_: The color, size, and position of the tick marks and grid lines on the axes.
 * _Legend_: The font, size, and position of the legend.

Resource:
 * https://mran.microsoft.com/snapshot/2017-02-04/web/packages/ggthemes/vignettes/ggthemes.html
 * https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/
 * https://ggplot2.tidyverse.org/reference/theme.html

```{r More on themes}
#Create a base plot to play with -- example of how you can change a plot just by changing theme
surface_temp_1999_plot <- PeterPaul.chem.nutrients %>% 
  filter(depth == 0 & year4 == 1999) %>% 
  ggplot(
    mapping = aes(
      x=sampledate,
      y=temperature_C,
      color=lakename)
    ) + 
  geom_line()

#Show the basic plot
surface_temp_1999_plot

#Apply a theme
surface_temp_1999_plot + 
  theme_gray()

#Apply a ggtheme
surface_temp_1999_plot + 
  theme_wsj() + 
  scale_color_wsj()

#Add a title
surface_temp_1999_plot + 
  ggtitle("Surface temperature, 1999") +
  labs(color='Lake') + 
  theme_wsj()
```


```{r Legend Elements}
#Create a custom theme - if you want to change a legend name, which is usually default, you can change that using color and labs...
my_theme <- theme_base() + 
  theme(
    line = element_line(
      color='red',
      linewidth =2
    ),
    legend.background = element_rect(
      color='grey',
      fill = 'orange'
    ),
    legend.title = element_text(
      color='blue'
    )
    
  )

#TIP: Expand the theme in the Environments

```


```{r useful theme creation template}
my_theme <- theme_base()  + 
  theme(
    #line =              element_line(),
    #rect =              element_rect(),
    #text =              element_text(),
    
    # Modified inheritance structure of text element
    #plot.title =        element_text(),
    #axis.title.x =      element_blank(),
    #axis.title.y =      element_blank(),
    #axis.text =         element_text(),
    
    # Modified inheritance structure of line element
    #axis.ticks =        element_line(),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    #plot.background =   element_rect(),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'right',
    
    #complete = TRUE
    )
```


### Scales
Scales in ggplot are used to map data values to aesthetic properties of plot elements, such as position, size, color, and shape. In addition to the basic types of scales, such as continuous and discrete scales, ggplot also provides date and time scales, and guide scales, which allow us to format and label the x-axis with appropriate date or time values, and to add legends and other visual guides to the plot. With a little bit of practice and experimentation, you can use ggplot to create beautiful and informative data visualizations with customized scales.

---

## Plotting continuous variables over time: Scatterplot and Line Plot

Exercise: build your own scatterplots of PeterPaul.chem.nutrients

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

# 1. 
# Go back to your PeterPaul lake dataset and create a ggPlot surface temperatures (x) by day of  year(x). Asking for SURFACE temperature -- need to filter your dataset. If you want surface temp, you need to filter by depth=0. you can do that prior to entering the ggplot environment or you can do it within.
# Color your points by year, and add facet by lake in two rows (each facet separated, using facet_wrap)
# Change the y axis lab name


#2. 
# Plot temperature by date. Now you don't need to filter your data set. Color your points by depth.
# Change the size of your point to 0.5 -- we have a lot of points to be added
# Change the color palette to magma and play with direction (+- 1), use scale color virides for example... which one makes more sense? are we going from light to dark or dark to light?
# Change x axis to include marker/labels every 5 years



```
## Plotting the relationship between two continuous variables: Scatterplot

Exercise: Plot relationships between air quality measurements

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

# 3. 
# Plot AQI values for ozone by PM2.5, colored by latitude 
# Make the points 50 % transparent
# Add a line of best fit for the linear regression of these variables.




```

## Plotting continuous vs. categorical variables

Exercise: Plot distributions of AQI values for EPAair.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

# 4. 
# Create several types of plots depicting PM2.5, divided by year. 
# Choose which plot displays the data best and justify your choice. 



```


## Visualization challenge

The following graph displays the counts of specific endpoints measured in neonicotinoid ecotoxicology studies. The way it is visualized, however, is not effective. Make the following coding changes to improve the graph: 

1. Change the ordering of the "Endpoint" factor (function: `reorder`) so that the highest counts are listed first (hint: FUN = length)
2. Plot the barplot with the reordered factor levels. Add this line of code to make the bars show up left to right:   scale_x_discrete(limits = rev(levels(Neonics$Endpoint)))
3. Adjust the x axis labels so they appear at a 45 degree angle. 
4. Change the color and/or border on the bars. Should you have a consistent color across all bars, or a different color for each bar?
```{r}

Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv")
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint))



```
