---
title: "8: Time Series"
author: "Environmental Data Analytics | Adapted by John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss the purpose and application of time series analysis for environmental data
2. Explore the components of times series: trend, seasonal, random
3. Learn how to perform stationarity test

## Set up

Today we will work with two datasets. The USGS dataset on discharge at the Eno River and a new dataset we haven't explored yet on wind speed. The data file is available at "./Data/Raw/Wind_Speed_PortArthurTX.csv". It contains average wind speed in monthly time steps (elevation = 5 meters). The data is available from NOAA National Centers for Environmental Information (NCEI) [here][https://www.ncdc.noaa.gov/cdo-web/datasets#GSOM].

```{r, message = FALSE}

library(tidyverse)
library(lubridate)
#install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

#Read Eno river data
EnoDischarge <- read.csv("./Data/Processed_KEY/USGS_Site02085000_Flow_Processed.csv",
                         stringsAsFactors = TRUE)
EnoDischarge$datetime <- as.Date(EnoDischarge$datetime, format = "%Y-%m-%d")

#Read wind speed data
wind_data <- read.csv(file="./Data/Raw/Wind_Speed_PortArthurTX.csv",header=TRUE,
                      stringsAsFactors = TRUE)
wind_data$DATE <- ym(wind_data$DATE)

```


## Time Series Analysis overview

Time series are a special class of dataset, where a response variable is tracked over time. The frequency of measurement and the timespan of the dataset can vary widely. At its most simple, a time series model includes an explanatory time component and a response variable. Mixed models can include additional explanatory variables (check out the `nlme` and `lme4` R packages). We will cover a few simple applications of time series analysis in these lessons, with references for how to take analyses further.

### Opportunities

Analysis of time series presents several opportunities. For environmental data, some of the most common questions we can answer with time series modeling are:

* Has there been an increasing or decreasing **trend** in the response variable over time?
* Can we **forecast** conditions in the future?

### Challenges

Time series datasets come with several caveats, which need to be addressed in order to effectively model the system. A few common challenges that arise (and can occur together within a single dataset) are: 

* **Autocorrelation**: Data points are not independent from one another (i.e., the measurement at a given time point is dependent on previous time point(s))
* **Data gaps**: Data are not collected at regular intervals, necessitating *interpolation* between measurements.
* **Seasonality**: seasonal patterns in variables occur at regular intervals, impeding clear interpretation of a monotonic (unidirectional) trend.
* **Heteroscedasticity**: The variance of the time series is not constant over time
* **Covariance**: the covariance of the time series is not constant over time

## Handling data gaps and missing data. Example: Eno River Discharge

River discharge is measured daily at the Eno River gage station. Since we are working with one location measured over time, this will make a great example dataset for time series analysis. 

Let's look at what the dataset contains for mean daily discharge.

```{r}
ggplot(EnoDischarge, aes(x = datetime, y = discharge.mean)) +
  geom_line() +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)"))
# missing data from 1971-1985
# a gap this large is an issue for TSA -- no good record to characterize any variability that might have happened over those years
# could split into early and late yrs and run the analysis separately
```

Notice there are missing data from 1971 to 1985. Gaps this large are generally an issue for time series analysis, as we don't have a continuous record of data or a good way to characterize any variability that happened over those years. We will illustrate a few workarounds to address these issues. 

Let's start by removing the NAs and splitting the dataset into the early and late years. 

```{r}
EnoDischarge.complete <- EnoDischarge %>%
  drop_na(discharge.mean)

#storing the earlier (before 1985) data
EnoDischarge.early <- EnoDischarge.complete %>%
  filter(datetime < as.Date("1985-01-01"))

#storing the later, post-1985 dataset
EnoDischarge.late <- EnoDischarge.complete %>%
  filter(datetime > as.Date("1985-01-01"))
```

## Decomposing a time series dataset

A given time series can be made up of several component series: 

1. A **seasonal** component, which repeats over a fixed known period (e.g., seasons of the year, months, days of the week, hour of the day)
2. A **trend** component, which quantifies the upward or downward progression over time. The trend component of a time series does not have to be monotonic.
3. An **error** or **random** component, which makes up the remainder of the time series after other components have been accounted for. This component reflects the noise in the dataset. (The part of the analysis that can't be explained by trend or seasonality and thus will be the hardest part to predict in our TS.)
4. (optional) A **cyclical** component, which repeats over periods greater than the seasonal component. A good example of this is El Niño Southern Oscillation (ENSO) cycles, which occur over a period of 2-8 years.

### Example: Eno discharge

We will decompose the EnoDischarge.late data frame for illustrative purposes today. It is possible to run time series analysis on detrended data by subtracting the trend component from the data. However, detrending must be done carefully, as many environmental data are bounded by zero but are not treated as such in a decomposition. If you plan to use decomposition to detrend your data, please consult time series analysis guides before proceeding.

We first need to turn the discharge data into a time series object in R. This is done using the `ts` function. Notice we can only specify one column of data and need to specify the period at which the data are sampled. The resulting time series object cannot be viewed like a regular data frame.

Note: time series objects must be equispaced. In our case, we have daily data with no NAs in the data frame, so we don't need to worry about this. We will cover how to address data that are not equispaced later in the lesson.

```{r}
EnoDischarge.late_ts <- ts(EnoDischarge.late$discharge.mean, start = c(1985,1), frequency = 365)
# alt: ts(EnoDischarge.late[[8]], start = c(1985,1), frequency = 365)
# when you define a TS object it's impt to define the starting point (what's your first observation? in our case, Jan 1985) -- 1985 is the yr, 1 is the month, and the frequency is 365 so the fxn knows you're working with daily data so it'll inc that by one day every time it's running a new obs
```

The `stl` function decomposes the time series object into its component parts. We must specify that the window for seasonal extraction is either "periodic" or a specific number of at least 7. The decomposition proceeds through a loess (locally estimated scatterplot smoothing) function.

```{r}
?stl
# the stl will perform a seasonal decomp of TS by loess (the method used to break the TS into 3 components)
# the first argument will always be the UNIVARIATE TS to be decomposed. so even if ur TS obj has more than 1 variable u need to run th decomp for each one separately. always only provide the decomp fxn with ONE VARIABLE
# s.window refers to the seasonal window -- u can define it as periodic. if u do it'll get whatever freq u specified or u can just use the span (in lags) of the seasonal component so for daily u could say s.window=365 or for monthly s.window=12. but if u specified that when u created the TS obj u can just say s.window=periodic.

# Generate the decomposition
EnoDischarge.late_Decomposed <- stl(EnoDischarge.late_ts, s.window = "periodic")

# Visualize the decomposed series. 
plot(EnoDischarge.late_Decomposed)
# first is the observed data
# second is the seasonal component
# third is trend component
# fourth is the random component (the remainder)
# it repeats itself every 365 observations

# We can extract the components and turn them into data frames
# creating a data frame 
EnoDischarge.late_Components <- as.data.frame(EnoDischarge.late_Decomposed$time.series[,1:3])
# perform some operations on the data frame
# change some columns -- add 2 more columns to components data frame
EnoDischarge.late_Components <- mutate(EnoDischarge.late_Components,
        Observed = EnoDischarge.late$discharge.mean,     
        Date = EnoDischarge.late$datetime)
# instead of 3 i have 5 variables in components data frame
# so that we can call ggplot

# Visualize how the trend maps onto the data
ggplot(EnoDischarge.late_Components) +
  # add a line for observed
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  # add a line for the trend component and add a special color so that the trend line pops out in comparison to observed data
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  # add geometric horizontal line at the intercept y=0 to call attention that we'll never have an obs that's less than 0 with environmental data, but once we generate a trend sometimes the trend component will be 0 so we need to be rly careful when we interpret these <0 observations
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))
# our observed data doesn't go below the dotted line

# Visualize how the seasonal cycle maps onto the data
# instead of plotting the trend, plot the seasonal component w the same color
ggplot(EnoDischarge.late_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Discharge (ft"^3*"/s)"))
# for the seasonal component it's clearly going below the y=0 line in every cycle of the seasonal component
```

Note that the decomposition can yield negative values when we apply a seasonal adjustment or a trend adjustment to the data. The decomposition is not constrained by a lower bound of zero as discharge is in real life. Make sure to interpret with caution!


## Trend analysis

Two types of trends may be present in our time series dataset: **monotonic/deterministic** or **stochastic**. Monotonic trends are a gradual shift over time that is consistent in direction, for example in response to land use change. 

A third type of trend we haven't talked about is the **step** trend, also knows as a level shift. Step trends are a distinct shift at a given time point, for example in response to a policy being enacted, or could be a change in the way the data is being recorded.


### Monotonic trend analysis

In general, detecting a monotonic trend requires a long sequence of data with few gaps. If we are working with monthly data, a time series of at least five years is recommended. Gaps can be accounted for, but a gap that makes up more than 1/3 of the sampling period is generally considered the threshold for considering a gap to be too long (a step trend analysis might be better in this situation). 

Adjusting the data may be necessary to fulfill the assumptions of a trend test. A common method to replace missing values is **interpolation**. Common interpolation methods: 

* **Piecewise constant**: also known as a "nearest neighbor" approach. Any missing data are assumed to be equal to the measurement made nearest to that date (could be earlier or later).
* **Linear**: could be thought of as a "connect the dots" approach. Any missing data are assumed to fall between the previous and next measurement, with a straight line drawn between the known points determining the values of the interpolated data on any given date.
* **Spline**: similar to a linear interpolation except that a quadratic function is used to interpolate rather than drawing a straight line.

#### Example: interpolation

The Eno River discharge data doesn't have any short periods of missing data, so interpolation would not be a good choice for that dataset. We will illustrate a linear interpolation using the wind speed dataset. 


```{r}
head(wind_data)
summary(wind_data$AWND)
# for the wind data we know that there's only 1, a single missing observation in the dataset. 
# in summary() we see there's only one not available observation

# Adding new column with no missing obs, just for illustration purpose
# In real applications you will simply replace NAs
wind_data_clean <- 
  wind_data %>% 
  # add a new column created using the zoo package and the na.approx fxn
  # na.approx fxn can be used to interpolate missing obs using either the linear or spline interpolation
  mutate( AWND.clean = zoo::na.approx(AWND) )

summary(wind_data_clean$AWND.clean)
#Note the NA is gone
# the mean, the median, and the max did not change -- the interpolation methods won't change the main properties of the distribution

ggplot(wind_data_clean ) +
  geom_line(aes(x = DATE, y = AWND.clean), color = "red") +
  geom_line(aes(x = DATE, y = AWND), color = "black") +
  ylab("Average Wind Speed") 
# this visualization shows the original plot with the missing data in black, and then the "clean" one with the interpolation in red to show where it was filled in and where you got a continuous data set -- NAs can't just be removed in TSA (unless they're just on the beginning or end of the data set)
```

### Monotonic trend analysis, continued

Specific tests for monotonic trend analysis are listed below, with assumptions and tips: 

* **linear regression**: no seasonality, fits the assumptions of a parametric test. Function: `lm`
* **Mann-Kendall**: no seasonality, non-parametric, missing data allowed. Function: `MannKendall()` (package: Kendall)
* **Seasonal Mann-Kendall**: seasonality, non-parametric `SeasonalMannKendall` (package: Kendall)
* **Spearman Rho**: no seasonality, non-parametric, missing data allowed. Function: `cor.test(method="spearman")` (package: stats)

You could also remove the seasonality of a data set and then run one of the tests that takes no seasonality

Specific test for stochastic trend analysis:

* **Augmented Dickey Fuller**: no seasonality, non-parametric, missing data not allowed. Function: `adf.test()` (package: tseries)


#### Example: monotonic trend analysis

Let's refer to our wind speed data. We already performed interpolation, but we still need to create our time series object and decompose the series to find out which stationarity test we can apply. 

Note that wind speed has a seasonal cycle. We might be interested in knowing how (if) speed has changed over the course of measurement while incorporating the seasonal component. In this case, we will use a Seasonal Mann-Kendall test to figure out whether a monotonic trend exists.


```{r}

# Generate time series (trend test needs ts, not data.frame)
# instead of manually providing the first yr and month of obs, i want it to be automatic
# the fxn "first" says i want my first month to be the equivalent month of my first obs in column DATE and f_year to be the first yr in the first obs for column DATE
f_month <- month(first(wind_data_clean$DATE))
f_year <- year(first(wind_data_clean$DATE))
# freq is 12 bc working w monthly data
wind_data_ts <- ts(wind_data_clean$AWND.clean,
                   start=c(f_year,f_month),
                   frequency=12) 


#decompose
wind_data_decomp <- stl(wind_data_ts,s.window = "periodic")
plot(wind_data_decomp)
# u can see the seasonal component is much easier to spot now (more stable and smoothe)
# can see trend going up and down a bit. can see overall decreasing trend but need to show it in numbers
# the remainder is the residual series

# Run SMK test
# can handle seasonal data
wind_data_trend1 <- Kendall::SeasonalMannKendall(wind_data_ts)

# Inspect results
wind_data_trend1
summary(wind_data_trend1)
# results from Mann-Kendall are pointing to p value < .05 and the nul hyp for the seasonal mann kendall ist hat the data is stationary so we can reject that null hyp and say we have a trend

wind_data_trend2 <- trend::smk.test(wind_data_ts)
# Inspect results
wind_data_trend2
summary(wind_data_trend2)
# the SMK test will show the results of the tau for each season of the year
# u can also see the value of S
# so u can see that for some seasons there's a stronger tendency of dec
# whereas seasons 4 and 5 the tendency of decrease is not that pronounced and their p values are much higher than .05 meaning for those two seasons it's actually stationary BUT the overall trend for the dataset is decreasing

#Visualization
wind_data_plot <-
ggplot(wind_data, aes(x = DATE, y = AWND)) +
  geom_point() +
  geom_line() +
  ylab("Average Wind Speed") +
  geom_smooth( method = lm )
print(wind_data_plot)
# negative slope shows decreasing trend for avg windspeed over time
```

What would we conclude based on these findings? 

> Answer:


## Forecasting with Autoregressive and Moving Average Models (ARMA)

We might be interested in characterizing a time series in order to understand what happened in the past and to effectively forecast into the future. Two common models that can approximate time series are **autoregressive** and **moving average** models. To classify these models, we use the  **ACF (autocorrelation function)** and the **PACF (partial autocorrelation function)**, which correspond to the autocorrelation of a series and the correlation of the residuals, respectively. the ACF and PACF will measure how many prev obs of ur data u need to be able to explain the variability of y.

**Autoregressive** models operate under the framework that a given measurements is correlated with  previous measurements. For example, an AR1 formulation dictates that a measurement is dependent on the previous measurement, and the value can be predicted by quantifying the lag. 

**Moving average** models operate under the framework that the covariance between a measurement and the previous measurement is zero. While AR models use past forecast *values* to predict future values, MA models use past forecast *errors* to predict future values.

Here are some great resources for examining ACF and PACF lags under different formulations of AR and MA models. 
https://nwfsc-timeseries.github.io/atsa-labs/sec-tslab-autoregressive-ar-models.html
https://nwfsc-timeseries.github.io/atsa-labs/sec-tslab-moving-average-ma-models.html

ARMA models require stationary data. This means that there is no monotonic trend over time and there is also equal variance and covariance across the time series. The function `adf.test` will determine whether our data are stationary. (If the data's non-stationary we need to perform differencing before fitting an ARMA model to the data set.) The null hypothesis is that the data are not stationary, so we infer that the data are stationary if the p-value is < 0.05.

While some processes might be easy to identify, it is often complicated to predict the order of AR and MA processes. To get around this issue, it is often necessary to run multiple potential formulations of the model and see which one results in the most parsimonious fit using AIC. The function `auto.arima` does this automatically. Put the data set ur trying to fit into the parenthesis and r will find all poss combos of order for autoreg and moving avg components and will use Akaike info criteria to determine what model is best fit for the data.

