---
title: "8: Lab - Time Series"
author: "Environmental Data Analytics | John Fay and Luana Lima"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M8
2. Explore the components of times series: trend, seasonal, random
3. Perform trend analysis on another dataset

## Set up

```{r, message = FALSE}

library(tidyverse)
library(lubridate)
#install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)
library(here)
here

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Import Datasets

Today we will work with data from the Climate Change Knowledge Portal from the World Bank Group. More specifically historical rainfall and temperature averages for Brazil. You will find two new data files on folder "/Data/Raw/". One with rainfall named "pr_1901_2016_BRA.csv" and another with temperature named "tas_1901_2016_BRA.csv". The data span the period from 1901 to 2016 in monthly steps. You can download the data [here][https://climateknowledgeportal.worldbank.org/download-data]

Research question: Can you see any changes on the rainfall regime or average temperature over time?

```{r}
#Exercise 1: Import both datasets using the read.csv function.
Rainfall_BR <- read.csv(here("Data/Raw/pr_1901_2016_BRA.csv"), stringsAsFactors = TRUE)
# when you loko at rainfall br, first column has rain info, 2nd has year, 3rd has month, but instead of just having the month abbrev it has a word avg so it wont be able to identify the month. you need to parse the column into one column by using the "separate" function. 

Temp_BR <- read.csv(here("Data/Raw/tas_1901_2016_BRA.csv"), stringsAsFactors = TRUE)

#Exercise 2: Tidy the rainfall data sets.
#a Rename the column with Rainfall to get rid of the dots.
#b Note that on both data sets that is a column with the month name and average. Convert it to a Month column by separating the string, i.e., month in one column and the word "Average" on another. Hint: use the separate() function in the pipe.
#c Now you should have a column with Month. Use the paste0() function to paste month and year together and create a date object.
#d Select only the columns of interest: Date and rainfall

Rainfall_BR_processed <-
  Rainfall_BR %>% 
  #rename first colum
  rename( Rainfall = Rainfall....MM.) %>% 
  #provide the fxn with the column that u want to be splitted and provide names for each column that will be created and the column separator, which is a space " " -- we need one more column before the month bc of an extra space BEFORE the month abbrev too so that's the first NULL column.
  separate(Statistics,c("Null","Month","Null2")," ") %>% 
  # need to put the month and year info together (column Year and month) using fxn "my" month-year to lubridate to transform into a date object. provide the combo of column Month - [dash] column Year
  mutate( Date = my(paste0(Month,"-",Year))) %>% 
  select(Date,Rainfall)

#Exercise 3: Repeat exercise 2 to the temperature dataset.
Temp_BR_processed <-
  Temp_BR %>% 
  rename( Temperature_C = Temperature....Celsius.) %>% 
  separate(Statistics,c("Null","Month","Null2")," ") %>% 
  mutate( Date = my(paste0(Month,"-",Year))) %>% 
  select(Date,Temperature_C)  


#Exercise 4: Join the temperature and rainfall into one tidy data frame with 3 columns: Date, Rainfall and Temperature. Doing an inner join where we're creating dataset that has complete info for each date. using inner_join instead of cbinding bc in time series we need to be rly careful with dates. if cbind, we'll never know if the rows are representing the same thing. inner join looks for columns of the same name and do a merge of the two datasets rather than just binding them. always use join rather than column bind to aavoid problem of completely missing an obs from one side or the other
BR_complete <- inner_join(Rainfall_BR_processed,Temp_BR_processed)
  
           
```


## Initial plots

```{r}

#Exercise 5: Check if there is any missing data for both temperature and rainfall series.
# since we did inner join, we can just look at the summary and if we had NA we'd have another column telling us how many NAs they found in the data. In the assignment, it's just one data source coming from one source so just using summary will not be enough. you will not be able to just run an na check bc the specific . SO CREATE A SEQUENCE USING SEQ FXN AND INSTEAD OF BEING A SEQUENCE OF NUMBERS U WILL DO SEQ OF DATES WHERE U START FROM FIRST DATE IN DATAFRAME USING FIRST() FXN ALL THE WAY DOWN TO LAST. SPECIFY FROM AND TO AND ALSO THE TIMESTEP. IF MONTHLY, 12 IS THE TIMESTEP. when u create these as a df what u will do is leftjoin the sequence of dates with ur data set. if u do a leftjoin it means u want all the rows in ur sequence to be represented in ur resulting df. what r will do if u have a month and yr with no records, r will populate that with NAs. so as soon as it does, when u run the summary u will be able to identify the NAs. this enables you to get to continuous data
summary(Rainfall_BR_processed$Rainfall)
# no NAs
summary(Rainfall_BR_processed$Date)
# no NAs
summary(Temp_BR_processed$Temperature_C)
# no NAs 
summary(Temp_BR_processed$Date)
# no NAs

#What we wanted:
summary(BR_complete$Rainfall)
summary(BR_complete$Temperature_C)

#Exercise 6: Plot both series over time on separate plots. Add a trend line using geom_smooth(). Can you say something about the trend just by visualizing the data?

ggplot(BR_complete, aes( x = Date, y = Rainfall)) +
  geom_line() +
  geom_smooth()
  
ggplot(BR_complete, aes( x = Date, y = Temperature_C)) +
  geom_line() +
  geom_smooth(method="lm")

# cannot see a trend for rainfall. the smoothed line doesn't rly have an inclination. BUT i can see a pattern; it's going up and down fast which means within a year so that's an indication that we prob have some seasonality happening in there. BR has wet summers and dry winters. you will see seasonality in precip in latin america
# Avg Temp: definitely see a trend, there's a clear upward trajectory of the trend line and we also can see seasonality. what this is telling us is that 
# how can we use time series to try to understand this a bit better? use TSA to prove what ur eyes are seeing 

```

## Creating time series objects

```{r}
#Exercise 7: Create a time series objects for each series using ts(). Make sure you specify the starting point and frequency.

# Generate time series for Rainfall
# instead of manually providing the first yr and month of obs, i want it to be automatic
# the fxn "first" says i want my first month to be the equivalent month of my first obs in column DATE and f_year to be the first yr in the first obs for column DATE
# common mistake when creating a ts is providing your entire data frame. whenever ur doing that, make sure u exclude the data column from the df otherwise the ts fxn will try to convert the date column to a ts too. 
f_month <- month(first(BR_complete$Date))
f_year <- year(first(BR_complete$Date))
# freq is 12 bc working w monthly data
BR_Rain_ts <- ts(BR_complete$Rainfall,
                 frequency=12,
                 start=c(1901,1))
BR_Temp_ts <- ts(BR_complete$Temperature_C,
                 frequency = 12,
                 start=c(f_year,f_month)) 

# BR_Rainfall_ts

#Temp_f_month <- month(first(Temp_BR_processed$Date))
#Temp_f_year <- year(first(Temp_BR_processed$Date))
# freq is 12 bc working w monthly data
#BR_Temperature_ts <- ts(Temp_BR_processed$Temperature_C,
                   #start=c(f_year,f_month),
                   #frequency=12) 
# BR_Temperature_ts

# alt how Luana did it: f_month <- month(first(BR_complete$Date))
# f_year <- month(first(BR_complete$Date))

#BR_Rain_ts <- ts(BR_complete$Rainfall, frequency = 12, start = c(1901,1))
#BR_Temp_ts <- ts(BR_complete$Temperature_C, frequency = 12, start=c(f_year,f_month))

# ts won't appear in environments as an obj. to see it. yr will be in rows and month will be in columns for date

```


## Decomposing a time series dataset

A given time series can be made up of several component series: 

1. A **seasonal** component, which repeats over a fixed known period (e.g., seasons of the year, months, days of the week, hour of the day)
2. A **trend** component, which quantifies the upward or downward progression over time. The trend component of a time series does not have to be monotonic.
3. An **error** or **random** component, which makes up the remainder of the time series after other components have been accounted for. This component reflects the noise in the dataset. 
4. (optional) A **cyclical** component, which repeats over periods greater than the seasonal component. A good example of this is El Niño Southern Oscillation (ENSO) cycles, which occur over a period of 2-8 years.


The `stl` function decomposes the time series object into its component parts. We must specify that the window for seasonal extraction is either "periodic" or a specific number of at least 7. The decomposition proceeds through a loess (locally estimated scatterplot smoothing) function.

```{r}

# Exercise 8: Use the stl() function to decompose both series. 
# Visualize the decomposed series using plot(). What components can you identify?

#decompose Rainfall
BR_Rainfall_Decomposed <- stl(BR_Rain_ts, s.window = "periodic")
plot(BR_Rain_decomp)
# u can see the seasonal component is much easier to spot now (more stable and smoothe)
# can see trend going up and down a bit. can see overall decreasing trend but need to show it in numbers
# the remainder is the residual series

#decompose Temperature
BR_Temp_decomp <- stl(BR_Temp_ts, s.window = "periodic")
# defaulting the freq to be the same freq i specified when defining the time series. r will not be able to fit a seasonal trend if it doesnt know whats ur seasonal lag
plot(BR_Temp_decomp)

# can identify data (original dataset), seasonality (seasonal component where u can see up and down movement a bit better than when u have entire dataset), trend (whatever movement not explained by the seasonality--overall some periods seem to be dec, some inc, which is prob why overall monotonic trend isn't v obvious. could be explained by e.g. rly dry season, effect of el niño / la niña), and remainder/random
# FOR RAINFALL: this is telling us that there are upward movement and downward movements, sometimes they cancel one another out so you could look at only the wet months and only the dry months and see if within themselves they have inc in intensity, if these overall analyses dont rly lead to strong conclusions. if u see a pattern in ur remainder/random or only have positive or negative numbers, something is wrong w ur data. u could be over- or under- feeding. u want the avg to be as close to 0 as possible and to remain aleatory 
# the grey bars on the right side of the plots show how big a role that component is playing in your data (e.g. here, the trend didn't have a big role bc it had to be zoomed into a lot/bigger bar). if they're the same size it means they're all playing the same-sized role in ur data set. here the remainder of the series is even more significant than the trend. could be an indication that thre's sth else we need to bring into the model.
# TEMP: saying seasonal and trend components are equally impt in estimating the temperature. an indication that the trend is statistically diff from 0. also giving additional info that the prev plot we had wasn't; when we look at overall upward monotonic trend we fail to look at how model has evolved over time. only a trend analysis can provide that the temps have rly increased over last 50 yrs -- the older data isn't rly representing where we think the future will be so we might have to throw out the data. 


# Exercise 9: Use the times series object generated with the stl() to extract the 
# components of each series and turn them into data frames.

# We can extract the components and turn them into data frames
# creating a data frame 
BR_Rain_Components <- as.data.frame(BR_Rainfall_Decomposed$time.series[,1:3])
# perform some operations on the data frame
# change some columns -- add 2 more columns to components data frame
BR_Temp_Components <- as.data.frame(BR_Temp_Decomposed$time.series[,1:3])

BR_Rain_Components <- mutate(BR_Rain_Components,
                             Observed = BR_complete$Rainfall,
                             Date = BR_complete$Date)
BR_Temp_Components <- mutate(BR_Temp_Components,
                             Observed = BR_complete$Temperature_C,
                             Date = BR_complete$Date)
# so that we can call ggplot

# Exercise 10: Visualize how the trend maps onto the data for both series
ggplot(BR_Rain_Components) +
  geom_line(aes(y = Observed, x = Date), size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  #geom_line(aes(y = seasonal, x = Date), color = "blue") +
  #geom_hline(yintercept = 0, lty = 2) +
  ylab("Rainfall")

ggplot(BR_Temp_Components) +
  geom_line(aes(y = Observed, x = Date), size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  ylab("Rainfall")




```


## Trend analysis

Monotonic trends are a gradual shift over time that is consistent in direction. Specific tests for monotonic trend analysis are listed below, with assumptions and tips: 

* **linear regression**: no seasonality, fits the assumptions of a parametric test. Function: `lm`
* **Mann-Kendall**: no seasonality, non-parametric, missing data allowed. Function: `Kendall::MannKendall()` or `trend::mk.test()`
* **Seasonal Mann-Kendall**: seasonality, non-parametric `Kendall::SeasonalMannKendall` or `trend::smk.test()`
* **Spearman Rho**: no seasonality, non-parametric, missing data allowed. Function: `stats::cor.test(method="spearman")`


```{r}

# Exercise 11: Apply one of the trend detection test to the original data set. Remember that the data has seasonal component, which test can be applied in this case?









# Exercise 12: Now remove the seasonality and apply another test. Compare the results with what you obtained on Exercise 9.











```

What would we conclude based on these findings? 

> Answer:


