---
title: "Assignment 3: Data Exploration"
author: "Gaby Czarniak"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Checking work directory,  message = FALSE}
getwd() # checking my working directory
print(getwd()) # printing (getwd()) so that TAs can see output
```

```{r Installing and loading packages}
# tinyverse and lubridate were already installed
# commenting out the install commands given that I need to knit and 
# including library command to load the packages

# install.packages("tidyverse") 
# install.packages("lubridate")

# library(tidyverse)
library(tidyverse)
# library(lubridate)
library(lubridate)
```

```{r Importing data}
# Uploading dataset on ecotoxicology
# Assigning name
# Reading strings in as factors
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = TRUE)
# Neonics

# Uploading and renaming dataset on litter and woody debris
# Assigning name
# Reading strings in as factors
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = TRUE)
# Litter
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: I did a brief internet search to learn more about neonicotinoids, and information at https://www.pnas.org/doi/10.1073/pnas.2017221117 described neonicotinoids as an exceptionally toxic class of insecticides, containing compounds that target specific receptors is insects' bodies. While intended to protect crops, these toxic substances are harming non-target species (including beneficial pollinators and arthropods that contribute to healthy soils) and traveling through food webs. They are threatening food webs and biodiversity. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris can affect carbon and nutrient cycling in forests. It can also affect soil moisture content, impacting the water available to forest plants, and create/maintain habitats for other organisms. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.  Litter, "defined as material that is dropped from the forest canopy and has a butt end diameter <2cm and a length <50 cm," is "collected in elevated 0.5m2 PVC traps.
 2. Woody debris, which they define as "material that is dropped from the forest canopy and has a butt end diameter <2cm and a length >50 cm," is collected in ground traps since it's longer and can't be collected in elevated traps. 
 3. Spatial sampling design indexes on the size of plots, whereas temporal sampling design indexes on ground traps that are sampled once per year.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dataset dimensions}
dim(Neonics)
print(dim(Neonics))

# The Neonics dataset has 4623 rows and 30 columns.
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Summary of commonly studied effects}
summary(Neonics$Effect)
print(summary(Neonics$Effect))
```

> Answer: Population is the most common of the effects studied (from "Effect" column), followed by mortality, behavior, feeding behavior, reproduction, and development--in that order. In the associated metadata file "ECOTOX_CodeAppendix.pdf", "Population" is defined as "measurements and endpoints relating to a group of organisms or plants of the same species occupying the same area at a given time." This effect is measured as abundance, so it makes sense that it would be commonly studied, as someone looking for patterns would be looking for data at a group level for a given species in a given time and place, and to draw conclusions at a population level, one would need many data entries. The next most common effect is mortality, which is defined as "measurements and endpoints where the cause of death is by direct action of the chemical." It makes sense to me that this would be commonly studied, because on the pyramid of severity of effect, with least severe at the bottom and increasing severity toward the top, mortality tops the pyramid. I can imagine someone interested in limiting severe effects of neonicotinoids would be looking to determine protection requirements near the most severe cases, assuming that those protections would be conservative in contributing to the protection of less severe cases, too. "Behavior" is defined as "overt activity of an organism represented by three effect groups - avoidance, general behavior, and feeding behavior. All measurements related to reproductive behavior are listed under the major effect group REP." Because behavior is an aggregate of three effects, it makes sense that it would appear commonly studied, especially that feeding behavior, itself, comes in as next-most-common. Reproduction is key to species survival, and development of an organism (defined in the metadata file as covering "toxicant effects on tissue organization in growing progeny") could affect reproductive capacity.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Summary of commonly studied species}
# Calling summary function to get first six most common species 
# by common name
summary(Neonics$Species.Common.Name, 6)
print(summary(Neonics$Species.Common.Name, 6))

# Because the sixth-most-common was "Other", 
# I also wanted to call the summary function using an argument 
# of 7, so that I could get a full six most common species 
# rather than five, plus "Other"
summary(Neonics$Species.Common.Name, 7)
print(summary(Neonics$Species.Common.Name, 7))

# sorting across all Species.Common.Name 
sort(summary(Neonics$Species.Common.Name))
# sorting across just the first seven Species.Common.Name
sort(summary(Neonics$Species.Common.Name, 7))
print(sort(summary(Neonics$Species.Common.Name, 7)))
```

> Answer: The six most commonly studied species in the dataset, by common name, are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. Although "Parasitic Wasp" was more than 58 entries less than "Honey Bee", I did notice that there seemed to be a typo in the Species.Common.Name entry of "Parastic Wasp", which could mean even more entries should be listed under "Parasitic Wasp" if it was, indeed, a typo. It seems like most of these species are important pollinators, which are key to our food chain. Parastic wasps also seem to have agricultural importance, due to their use as biological control agents against pests, according to https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4516919/.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Getting class of column data}
# Asking r to tell me the class of the column's data
class(Neonics$Conc.1..Author.) 
print (class(Neonics$Conc.1..Author.))
```

> Answer: The Conc.1..Author. column in the dataset is a factor. It's not numeric because I asked r to read the string in as factors. Storing the data as factors rather than numeric variables allows us to do more and different things with the data. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Generating frequency line graph, fig.height = 3, fig.width = 4}
# Plotting a frequency line graph of the number of studies (y axis) 
# by Publication Year (x axis)
#
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50) 

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Comparing studies from test locations, fig.height = 3, fig.width = 4}
# Plotting a frequency line graph of the number of studies (y axis) 
# by Publication Year (x axis)
# 
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50)

# Color-coding the count of number of studies by their Test Location
#
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) +
  theme(legend.position = "top")

# Confirming most common test locations
sort(summary(Neonics$Test.Location))
print(sort(summary(Neonics$Test.Location)))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Overall, the most common test locations are lab locations, followed by field natural, then field artifical, then field undeterminable. They do differ over time--lab locations have definitely been more common in the 2010s, but there have been moments in the mid-to-late 1990s where field natural were the most common test locations, as well as in 2009, when they were more common than lab locations.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Creating Endpoint bar graph, fig.height = 3, fig.width = 4}
# Plotting Endpoint counts as bar graph
ggplot(Neonics, aes(x = Endpoint)) +
  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  #may be an error with theme because when I run it, none of my data is visible
  geom_bar()

print(ggplot(Neonics, aes(x = Endpoint)) +
  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  #may be an error with theme because when I run it, none of my data is visible
  geom_bar())
```

> Answer: NOEL and LOEL are the two most common end points. They are defined in ECOTOX_CodeAppendix as: for LOEL, "Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls (LOEAL/LOEC)"), and for NOEL, "No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test (NOEAL/NOEC)" -- both for Terrestrial database usage.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Determining data class}
# Checking class of collectDate in Litter
# class tells me that it is a factor
class(Litter$collectDate)
print(class(Litter$collectDate))

# I double check that it's not a date
is.Date(Litter$collectDate)
print(is.Date(Litter$collectDate))

#date_obj_collectDate <- (Litter$collectDate)
#format(date_obj_collectDate, format = "%y - %m - %d")
#date_obj_collectDate <- ymd(Litter$collectDate)
date_obj_collectDate <- as.Date(Litter$collectDate, format = "%y-%m-%d") 
date_obj_collectDate
is.Date(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Determining number of plots sampled at Niwot Ridge}
# Removing duplicates from plotID
unique(Litter$plotID)
print(unique(Litter$plotID))

```

> Answer: Twelve plots were sampled at Niwot Ridge. The unique function removes duplicates from plotID. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Creating functionalGroup bar graph, fig.height = 3, fig.width = 4}
# Plotting functionalGroup counts as bar graph
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar()

print(ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar())
  
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Creating box and violin plots, fig.height = 3, fig.width = 4}
#
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))
print(ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)))

#
ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75))
print(ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75)))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The dryMass is being shown by functionalGroup and functionalGroup is made up of elements like flowers and leaves, whose masses don't have associated probability densities.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles, Mixed litter, and Twigs/branches tend to have the highest biomass at these sites.
